### 2021-06-14 JSS/using_drtmle_jss.Rmd
touch-ups up through CIs

This section contains some more advanced mathematical material and may be
omitted by users interested only in implementation. For simplicity, we focus
this discussion on the standard TMLE/AIPTW estimators. The ideas generalize
immediately to the DRTMLE estimator, but the formulas are more complex.

Reported standard error estimates are based on an estimate the asymptotic
variance of the estimators. This variance can be characterized by the estimators
influence function.  For example, let $\psi_{n}(1)$ be an AIPTW estimate of
$\psi_0(1)$. Under regularity conditions,
\begin{equation*}
  \psi_{n}(1) - \psi_0(1) = \frac{1}{n} \sum_{i=1}^n
    D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O_i) + o_{p}(n^{-1/2}) \ ,
\end{equation*}
where
\begin{equation*}
  D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O_i) = \frac{I(A_i = 1)}
  {g_0(1 \mid W_i)} \{ Y_i - \bar{Q}_0(1, W_i)\} + \bar{Q}_0(1, W_i) -
  \int \bar{Q}_0(1, w) dQ_{W,0}(w) \ .
\end{equation*}
Thus, by the central limit theorem, $n^{1/2} \psi_{n}(1)$ converges in
distribution to a Normal distribution with mean $\psi_0(1)$ and variance
\begin{equation*}
  \sigma^2_0 = E_0( [ D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O) -
  E_0\{D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O)\} ]^2 ) \ .
\end{equation*}
An estimate of $\sigma^2_0$ is naturally obtained by taking the empirical
variance of $D_i = D^{\star}(\bar{Q}_n, g_n, Q_{W,n})(O_i)$,
\begin{equation*}
  \sigma^2_n = \frac{1}{n} \sum_{i=1}^n \left\{ D_i - \frac{1}{n} \sum_{j=1}^n
  D_j \right\}^2 \ .
\end{equation*}
Thus, a natural estimate of the standard error of $\psi_{n}(1)$ is
$\sigma_n / n^{1/2}$. This idea generalizes naturally to estimating the
asymptotic covariance matrix of a vector $n^{1/2} (\psi_{n}(a) : a)$.

@@

For TMLE-based estimators, it is standard practice to construct $D_i$ above using the *targeted* 
version of the OR (denoted by $\bar{Q}_n^*$ above). Recall that such estimates are obtained by modifying
an initial OR estimate in such a way as to ensure a particular (set of) equation(s) is (are) solved.
However, simulations have shown that better standard error estimates for TMLE may be achieved by instead
using the *initial* OR estimate in the computation of standard errors, as described in the previous
subsection. The `drtmle` function includes the `targeted_se` option to this end.

@@

In simulations, we often find that when employing machine learning to estimate the OR and/or PS, 
the estimated standard errors are *too small* in small to moderate-sized samples, leading to 
under-coverage of confidence intervals and possibly too-large type I errors of hypothesis tests that
are based on these standard error estimates. This should not be too surprising when studying the form
of the standard error estimates. In particular, they will include a term that looks like 
a (weighted) mean squared-error (MSE) of the OR. Because machine learning algorithms often set out 
to minimize empirical MSE as part of their training, we often find that the empirical MSE is biased 
too small relative to the true MSE. Thus, in order to more accurately estimate MSE of a machine learning
algorithm, we typically rely on cross-validation. The same trick can be used here to obtain a more
conservative standard error estimate.

Suppose we use $V$-fold cross-validation to obtained $V$ training-sample-specific estimates of the OR,
 $\bar{Q}_{n,v}, v = 1, \dots, V$ and the PS, $g_{n,v}, v = 1, \dots, V$. For $i = 1, \dots, n$, let $\bar{Q}_{n,i}^{cv}(1)$ denote the estimate of $\bar{Q}_{0}(a, W_i)$ obtained when observation $i$ was in the validation fold. That is, $\bar{Q}_{n,i}^{cv}$ is an out-of-sample prediction of $Y_i$. We define
 $g_{n,i}^{cv}$ similarly. Using these estimates, we can define \[
 D_i^{cv} = \frac{I(A_i = 1)}{g_{n,i}^{cv}} \{ Y_i - \bar{Q}_{n,i}^{cv}(1)\} + \bar{Q}_{n,i}^{cv}(1) - \frac{1}{n} \sum_{i=1}^n \bar{Q}_{n,i}^{cv}(1) 
 \]
and compute an estimate of $\sigma^2_0$ using \[
  \sigma^2_{n,cv} = \frac{1}{n} \sum_{i=1}^n \left\{ D_i^{cv} - \frac{1}{n} \sum_{j=1}^n D_j^{cv} \right\}^2 \ . 
\]

Cross-validated standard errors can be obtained by setting `se_cv = "full"` and `se_cvFolds` to a number greater than 1.

@@

Obtaining fully cross-validated standard errors can be fairly time intensive when super learner is
used to estimate the OR and PS. In `drtmle` we give an option to compute standard error estimates that
are *partially cross-validated* and do not require any additional fitting beyond a standard super learner. 
Here we describe briefly how these estimates are obtained and show calls to `drtmle` to obtain these estimates.

Suppose we have $K$ candidate regressions in our super learner library for both the OR and PS 
(in practice, the number could be different for the OR as for the PS). Recall that a super learner
estimate of for example $\bar{Q}_0(a, w)$ is obtained as a weighted combination of estimates 
from each candidate regression. Let $\bar{Q}_{n,k}$ denote the estimate of $\bar{Q}_{0}$ obtained 
by training algorithm $k$ using the full data set. The super learner estimate of $\bar{Q}_{0}(a,w)$ is \[
  \bar{Q}_{n,SL}(a, w) = \sum_{k = 1}^K \alpha_{n,k} \bar{Q}_{n,k}(a, w) \ , 
\]
where $\alpha_n = (\alpha_{n,1}, \dots, \alpha_{n,K})$ is a weight vector that is selected by minimizing
a cross-validated risk criteria. That is, each of the $K$ algorithms is trained according to a $V$-fold
cross-validation scheme, resulting in $V + 1$ fits of each algorithm (one in each of $V$ training folds
plus one obtained by fitting the algorithm using the full data). Let $\bar{Q}_{n,k,v}$ denote the $k$-th
algorithm trained using the $v$-th training fold. For $i = 1,\dots,n$, let $\bar{Q}_{n,k,i}^{cv}(1)$ 
denote the estimate of $\bar{Q}_0(1, W_i)$ implied by the fitted algorithm $\bar{Q}_{n,k,v}$, which 
was obtained when observation $i$ was in the validation sample. That is, for observations with
$A_i = 1$, $\bar{Q}_{n,k,i}^{cv}(1)$ is an out of sample prediction of $Y_i$ based on the
$k$-th algorithm. Now, we can construct a *partially cross-validated* super learner prediction as follows. For $i = 1, \dots, n$ let \[
  \bar{Q}_{n,SL,i}^{cv}(1) = \sum_{k = 1}^K \alpha_{n,k} \bar{Q}_{n,k,i}^{cv}(1)
\]
be the estimate obtained by ensembling the training-set-specific algorithm fits based on $\alpha_n$. 
We can use these estimates to produce our *partially cross-validated* standard error estimate as follows. We define \[
 D_i^{pcv} = \frac{I(A_i = 1)}{g_{n,SL,i}^{cv}} \{ Y_i - \bar{Q}_{n,SL,i}^{cv}(1)\} + \bar{Q}_{n,SL,i}^{cv}(1) - \frac{1}{n} \sum_{i=1}^n \bar{Q}_{n,SL,i}^{cv}(1) 
 \]
and compute an estimate of $\sigma^2_0$ using \[
  \sigma^2_{n,pcv} = \frac{1}{n} \sum_{i=1}^n \left\{ D_i^{pcv} - \frac{1}{n} \sum_{j=1}^n D_j^{pcv} \right\}^2 \ . 
\]

Note that is not a proper cross-validated estimate, since observation $i$ is used to compute the 
weight vector $\alpha_n$. However, such estimates can be obtained based on the output of a single 
round of super learning, thereby providing improved standard error estimates at minimal additional
computational cost.

Here, we demonstrate how to obtain such estimates using `drtmle`.

@@

Planned extensions of the `drtmle` package include extensions of longitudinal
settings involving treatments at multiple time points and inclusion of the
collaborative targeted maximum likelihood estimator (C--TMLE) of
\citet{vdl:2014:ijb} that also facilitates collaborative doubly-robust
inference.

>>>>

This section contains some more advanced mathematical material and may be
skipped over by users interested only in implementation. For simplicity, we focus
this discussion on the standard TML/AIPTW estimators. The ideas generalize
immediately to the DR-TML estimator, but the formulas are more complex in this
latter case.

Reported standard error estimates are based on an estimate of the asymptotic
variance of the estimators. This variance can be characterized by influence
function of the estimators. For example, let $\psi_{n}(1)$ be an AIPTW estimate
of $\psi_0(1)$. Under regularity conditions,
\begin{equation*}
  \psi_{n}(1) - \psi_0(1) = \frac{1}{n} \sum_{i=1}^n
    D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O_i) + o_{p}(n^{-1/2}) \ ,
\end{equation*}
where
\begin{equation*}
  D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O_i) = \frac{I(A_i = 1)}
  {g_0(1 \mid W_i)} \{ Y_i - \bar{Q}_0(1, W_i)\} + \bar{Q}_0(1, W_i) -
  \int \bar{Q}_0(1, w) dQ_{W,0}(w) \ .
\end{equation*}
Thus, by the central limit theorem, $n^{1/2} \psi_{n}(1)$ converges in
distribution to a normal distribution with mean $\psi_0(1)$ and variance
\begin{equation*}
  \sigma^2_0 = E_0( [ D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O) -
  E_0\{D^{\star}(\bar{Q}_0, g_0, Q_{W,0})(O)\} ]^2 ) \ .
\end{equation*}
An estimate of $\sigma^2_0$ is naturally obtained by taking the empirical
variance of $D_i = D^{\star}(\bar{Q}_n, g_n, Q_{W,n})(O_i)$, that is,
\begin{equation*}
  \sigma^2_n = \frac{1}{n} \sum_{i=1}^n \left\{D_i - \frac{1}{n} \sum_{j=1}^n
  D_j \right\}^2 \ .
\end{equation*}
Thus, a natural estimate of the standard error of $\psi_{n}(1)$ is $\sigma_n /
n^{1/2}$. Conveniently, this idea generalizes naturally to estimating the
asymptotic covariance matrix of a vector $n^{1/2} (\psi_{n}(a) : a)$.

@@

For TML estimators, it is standard practice to construct $D_i$ using the
*targeted* version of the OR (denoted by $\bar{Q}_n^{\star}$ above). Recall that
such estimates are obtained by modifying an initial OR estimate so as to ensure
a particular (set of) equation(s) is (are) solved.  However, simulations have
shown that better standard error estimates for TMLE may be achieved by instead
using the *initial* OR estimate in the computation of standard errors, as
described in the preceding subsection. To this end, The `drtmle` function
includes the `targeted_se` option.

@@

Simulation studies have revealed that, when employing machine learning to
estimate the OR and/or PS, the estimated standard errors are often *too small*,
in small to moderate-sized samples. Problematically, this leads directly to
under-coverage of confidence intervals and possibly inflated Type-I error rates
of hypothesis tests based on these standard error estimates. This should not be
too surprising, as the phenomenon is related to the form of the standard error
estimates. In particular, such estimates include a term that resembles a
(weighted) mean squared error (MSE) of the OR. Since machine learning algorithms
often set out to minimize empirical MSE as part of their training process, it
follows that the empirical MSE often ends up biased towards zero relative to the
true MSE. To more accurately estimate the MSE of a machine learning algorithm,
cross-validation can be used. The same sample-splitting idea can be used to
obtain more conservative standard error estimates.

Suppose we use $V$-fold cross-validation to obtaine $V$ training-sample-specific
estimates of the OR, $\bar{Q}_{n,v}, v = 1, \dots, V$ and the PS, $g_{n,v},
v = 1, \dots, V$. For $i = 1, \dots, n$, let $\bar{Q}_{n,i}^{cv}(1)$ denote the
estimate of $\bar{Q}_{0}(a, W_i)$ obtained when observation $i$ was in the
validation fold. That is, $\bar{Q}_{n,i}^{cv}$ is an out-of-sample prediction of
$Y_i$. We define $g_{n,i}^{cv}$ similarly. Using these estimates, we can define
\begin{equation*}
 D_i^{cv} = \frac{I(A_i = 1)}{g_{n,i}^{cv}} \{ Y_i - \bar{Q}_{n,i}^{cv}(1)\} +
 \bar{Q}_{n,i}^{cv}(1) - \frac{1}{n} \sum_{i=1}^n \bar{Q}_{n,i}^{cv}(1)
\end{equation*}
and compute an estimate of $\sigma^2_0$ using
\begin{equation*}
  \sigma^2_{n,cv} = \frac{1}{n} \sum_{i=1}^n \left\{ D_i^{cv} -
  \frac{1}{n} \sum_{j=1}^n D_j^{cv} \right\}^2 \ .
\end{equation*}

In `drtmle`, cross-validated standard error estimates can be obtained by setting
`se_cv = "full"` and `se_cvFolds` to a number greater than 1. For example,

@@

Obtaining fully cross-validated standard errors can be fairly time-intensive
when super learner is used to estimate the OR and PS. In `drtmle`, we give an
option to compute standard error estimates that are *partially cross-validated*
and do not require any additional fitting beyond a standard super learner. Here,
we describe briefly how these estimates are obtained and show calls to `drtmle`
for obtaining such estimates.

Suppose we have $K$ candidate regressions in our super learner library for both
the OR and PS (in practice, the number could be different for the OR and for the
PS). Recall that a super learner estimate of, for example, $\bar{Q}_0(a, w)$ is
obtained as a weighted combination of estimates from each candidate regression.
Let $\bar{Q}_{n,k}$ denote the estimate of $\bar{Q}_{0}$ obtained by training
algorithm $k$ using the full data set. The super learner estimate of
$\bar{Q}_{0}(a,w)$ is
\begin{equation*}
  \bar{Q}_{n,SL}(a, w) = \sum_{k = 1}^K \alpha_{n,k} \bar{Q}_{n,k}(a, w) \ ,
\end{equation*}
where $\alpha_n = (\alpha_{n,1}, \dots, \alpha_{n,K})$ is a weight vector that
is selected by minimizing a cross-validated risk criterion. That is, each of the
$K$ algorithms is trained according to a $V$-fold cross-validation scheme,
resulting in $V + 1$ fits of each algorithm (one in each of $V$ training folds,
plus one obtained by fitting the algorithm using the full data). Let
$\bar{Q}_{n,k,v}$ denote the $k$\textsuperscript{th} algorithm trained using the
$v$\textsuperscript{th} training fold. For $i = 1, \dots, n$, let
$\bar{Q}_{n,k,i}^{cv}(1)$ denote the estimate of $\bar{Q}_0(1, W_i)$ implied by
the fitted algorithm $\bar{Q}_{n,k,v}$, which was obtained when observation $i$
was in the validation sample. That is, for observations with $A_i = 1$,
$\bar{Q}_{n,k,i}^{cv}(1)$ is an out-of-sample prediction of $Y_i$ based on the
$k$\textsuperscript{th} algorithm. Now, we can construct a *partially
cross-validated* super learner prediction as follows. For $i = 1, \dots, n$ let
\begin{equation*}
  \bar{Q}_{n,SL,i}^{cv}(1) = \sum_{k = 1}^K \alpha_{n,k}
    \bar{Q}_{n,k,i}^{cv}(1)
\end{equation*}
be the estimate obtained by ensembling the training-set-specific algorithm fits
based on $\alpha_n$. We can use these estimates to produce our *partially
cross-validated* standard error estimate as follows. First, define
\begin{equation*}
  D_i^{pcv} = \frac{I(A_i = 1)}{g_{n,SL,i}^{cv}} \{ Y_i -
    \bar{Q}_{n,SL,i}^{cv}(1)\} + \bar{Q}_{n,SL,i}^{cv}(1) -
    \frac{1}{n} \sum_{i=1}^n \bar{Q}_{n,SL,i}^{cv}(1),
\end{equation*}
then, compute an estimate of $\sigma^2_0$ using
\begin{equation*}
  \sigma^2_{n,pcv} = \frac{1}{n} \sum_{i=1}^n \left\{ D_i^{pcv} -
  \frac{1}{n} \sum_{j=1}^n D_j^{pcv} \right\}^2 \ .
\end{equation*}

Note that is not a properly cross-validated estimate, since observation $i$ is
used to compute the weight vector $\alpha_n$. However, such estimates can be
obtained based on the output of a single round of super learning, thereby
providing improved standard error estimates at minimal additional computational
cost.

Below, we demonstrate how to obtain such estimates using `drtmle`.

@@

Planned extensions of the `drtmle` package include extensions of longitudinal
settings involving treatments at multiple time points and inclusion of the
collaborative targeted maximum likelihood estimator (C-TMLE) of
\citet{vdl:2014:ijb} that also facilitates collaborative doubly-robust
inference.
